Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-14 application.tex
updating

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for three
opacity scales: $\hat{\kappa}_d=0.1, \,1 $ (MMSN) and $10$ at two
heights: $z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in each
panel.

@@

In fact, for $\khat \lesssim 30$ we either only find the
fundamental mode, or that it is the most unstable, and the associated
growth rates are insensitive to $z_\mathrm{max}$. This is consistent with our
analytical discussion (\S\ref{iso_vsi_beta_crit}) and previous
numerical results (\S\ref{therm_relax_eff}), which indicate  
higher order modes are more easily stabilized by finite thermal
relaxation.

However, for large $\khat$ and/or $z_\mathrm{max}$, we also found
surface modes with shorter growth times than those shown in
Fig. \ref{mmsn_overall}. Such modes may not be robust because    
they are affected by boundary conditions (\S\ref{surf_comment}) and/or
viscosity in a real disk.

@@

\subsection{Comparing the VSI in the inner and outer parts of the MMSN} 
We compare the VSI in the MMSN at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with much
reduced growth rates (when scaled by the local 
rotation).

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column due to the vertically
global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests that the VSI can operate 
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer at $r=5$AU than at $r=50$AU: %  for which  $\beta <
% \beta_\mathrm{crit}  \forall z$: 
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

>>>>

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for opacity scales
$\hat{\kappa}_d=0.1, \,1 $ (MMSN) and $10$, at $z=0$ and $z=3H$. We
also plot Eq. \ref{bcrit_mmsn} in each panel.

@@

In fact, we either only find the fundamental mode, or that it is the
most unstable, except for the appearence of some surface modes when considering large
$\khat\gtrsim 50$. However, surface modes are dependent on
boundary conditions and disturbances with large $\khat$ are more likely
subject to viscous decay, as discussed above. The dominance of the
fundamental mode is also observed in hydrodynamic
simulations of \cite{stoll14} which treat radiative cooling properly.

% The dominance of the
% fundamental mode is consistent with our analytical discussion
% (\S\ref{iso_vsi_beta_crit}) and previous numerical results (\S\ref{therm_relax_eff}), which indicate  
% higher order modes are more easily stabilized by finite thermal
% relaxation.

% In fact, for $\khat \lesssim 30$ we either only find the
% fundamental mode, or that it is the most unstable, and the associated
% growth rates are insensitive to $z_\mathrm{max}$. This is consistent with our
% analytical discussion (\S\ref{iso_vsi_beta_crit}) and previous
% numerical results (\S\ref{therm_relax_eff}), which indicate  
% higher order modes are more easily stabilized by finite thermal
% relaxation.   
% However, for large $\khat$ and/or $z_\mathrm{max}$, we also found
% surface modes with shorter growth times than those shown in
% Fig. \ref{mmsn_overall}. Such modes may not be robust because    
% they are affected by boundary conditions (\S\ref{surf_comment}) and/or
% viscosity in a real disk.

@@

\subsection{Numerical examples} 
We compare the VSI in the MMSN at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

@@

The outer disk ($r=50$AU) is optically thin,
$\beta<\beta_\mathrm{crit}$  and  does not depend on height. 
Accordingly we expect, and find, the VSI 
operates at $r=50$AU. The inner disk ($r=5$AU) is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with much
reduced growth rates (when scaled by the local 
rotation).

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column due to the vertically
global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests that the instability can operate 
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer at $r=5$AU than at $r=50$AU: %  for which  $\beta <
% \beta_\mathrm{crit}  \forall z$: 
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

### 2015-03-14 numerical.tex
updating

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. Note that for the standard disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$.  
Thus, tracking the fundamental mode is a more robust way of 
assessing stability as $\beta$ is increased from zero, rather than
searching for the most unstable mode, which are affected by boundary 
conditions.

@@

Fig. \ref{gcorr_compare_vnoniso} plots the fundamental VSI growth
rates as a function of $\beta$ for $\gamma\in[1.4,2.5]$. In agreement
with \citeauthor{nelson13}, in  
the neutrally-stratified case $\gamma=\Gamma$ the disk is unstable 
even for $\beta\gg 1$. (Note that in the adiabatic limit 
this disk can be unstable according to the
Solberg-Hoiland criterion, Eq. \ref{solberg2}.)

>>>>

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. Note that for the fiducial disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$.  
Thus, tracking the fundamental mode is a more robust way of 
assessing stability as $\beta$ is increased from zero, rather than
searching for the most unstable mode, which may correspond to modes
that depend on boundary conditions.

@@

Fig. \ref{gcorr_compare_vnoniso} plots the fundamental VSI growth
rates as a function of $\beta$ for $\gamma\in[1.4,2.5]$. In agreement
with \citeauthor{nelson13}, in   
the neutrally-stratified case $\gamma=\Gamma$ the disk is unstable 
even for $\beta\gg 1$. This is due to the absence of a stabilizing
vertical entropy gradient ($N_z^2\equiv 0$). % (Note that in the adiabatic limit 
% this disk can be unstable according to the
% Solberg-Hoiland criterion, Eq. \ref{solberg2}.)

### 2015-03-14 summary.tex
updating

\section{Summary and discussion}\label{summary}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

% The VSI was analyzed earlier in 
% the local limit by \cite{urpin98} and 
% \cite{urpin03} and simulated by \cite{arlt04}. Recent non-linear
% numerical simulations by \cite{nelson13} and \cite{stoll14} indicate
% the VSI generally requires a short thermal timescale.

We generalized previous vertically global linear stability calculations
of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, by relaxing the Boussinesq and vertically local
approximations adopted in that analysis.

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach to model thermal relaxation was also
employed in the global numerical simulations of \cite{nelson13}, and
allows the thermodynamic response of the disk to be controlled by
varying the parameter $\beta$. However, it is possible to model
$\beta$ self-consistently by relating it to thermal timescales
expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for non-isothermal
perturbations provided $\beta\lesssim1$.  
This is convenient because rapid thermal relaxation ($\beta\ll1$) is
generally required for the VSI.  However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$), since
this is the relevant case for protoplanetary disks that are passively irradiated by its central star 
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation, but in the limit of low-frequency perturbations in
nearly-Keplerian disks. These approximations were made for simplicity,
but also motivated by recent numerical simulations
\citep{nelson13}. We obtained explicit solutions for the VSI with
thermal relaxation in the thin disk limit.

% For isothermal perturbations ($\beta=0$) we explicitly
% demonstrated instability  associated with vertical shear, and that the 
% VSI growth rate is limited by the maximum vertical shear within the 
% vertical domain of interest. 
% For perturbations with finite thermal relaxation, the general
% dispersion relation (Eq. \ref{relax_disp}) is quite
% complicated. However, for the fundamental VSI some simplification is
% possible. In particular,

The fundamental VSI mode is characterized by a symmetric vertical
velocity perturbation with zero nodes and $\delta
v_z\sim\mathrm{constant}$ near the mid-plane, and by an anti-symmetric
density perturbation with one node at $z=0$.  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be the dominant mode,
 and it is robust against boundary conditions. The fundamental mode is
 thus likely the most relevant VSI mode in protoplanetary disks.

We quantified the critical thermal relaxation timescale 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the fundamental VSI
operates. %  We expect $\beta<\beta_\mathrm{crit}$ to be a sufficient
% condition for the VSI to operate.
We have therefore provided a quantitative measure for the `fast'
thermal relaxation requirement for the VSI in astrophysical disks.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analysis, and accounted for
the background radial disk structure for consistency.  Our
numerical solutions are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.  %for regimes of interest

% as given
% by Eq. \ref{iso_vsi_cond}, provides a good measure of the required
% th 
%  except for perturbations with small radial
% wavenumber ($\khat\to0$) where the low-frequency approximation is
% inappropriate. The radially local model is also not valid for such modes.

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks, using a model for the the Minimum Mass Solar Nebulae
(MMSN) described in \cite{chiang10}. We  
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity, which results in a $\beta$ that
depends on both the perturbation lengthscale as well as the height
from the midplane. We find the MMSN is unstable to the VSI
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating $\beta <
\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN.  This simple criteron may
be a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

%  However, we note the strong
% temperature dependence of the physical thermal relaxation timescale
% in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
% radial scale). Thus, increasing the disk temperature even slightly can significantly 
% shorten the typical growth times for the VSI. This is also attributed
% to a larger value of $\beta_\mathrm{crit}\propto\epsilon$, since increasing $T$ also
% increases the aspect-ratio $\epsilon$.

%relation with kley simulations

@@

The convective overstability corresponds to growing epicyclic motions.
It is most effective for disturbances with $|k_x|\ll |k_z|$, where
$k_z$ is a vertical wavenumber, and $\beta\sim 1$. These conditions
are mutually exclusive with the VSI. 
The maximum growth 
rate for convective overstability is given by 
\begin{align}
  \nu_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.

Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
we find  
\begin{align}
  \nu_\mathrm{conv,max} =
  -\frac{\epsilon^2\Omega_k}{4\gamma\Gamma}\left(p+q\right)
  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
\end{align}
% We see that convective overstability is generally slower 
% than the VSI because $\nu_\mathrm{conv,max} = O(\epsilon^2\Omega_k)$,
% whereas the maximum VSI growth rate is $O(\epsilon\Omega_k)$. 
For the MMSN considered in this work (\S\ref{application} and Appendix
\ref{mmsn}), convective overstability does not apply because
$N_r^2>0$.

However, for more general disk structures convective overstability is
possible. In fact, the models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$. %show early result
We will explore convective overstability more fully in a follow-up
study.  Although our linear code can be used for convective
overstability in principle, the fact convective overstability is
relevant for perturbations with $k_x\to0$ suggest radially global disk
models may be more appropriate.

% Nevertheless, it will be important to expand the parameter space covered in
% this study to include convective overstability,

%self-consistence may require radially global disk models

\subsubsection{Improved thermal evolution.} 
We assumed non-adiabatic effects
can be represented by a thermal relaxation timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties, but one can improve the realism further by generalizing the
opacity model adopted here to be density-dependent,  i.e. $\kappa_d =
\kappa_d(\rho, T)$. 
%two temperature
% structure

However, going beyond the framework of thermal
relaxation requires a proper treatment for radiative diffusion, which
involves the Lapacian operator and additional boundary conditions
must be supplemented. We will consider these improvements in a
follow-up study.

\subsubsection{Additional physics.}

Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a protoplanetary disk, where we estimate the VSI 
operates, disk self-gravity (SG) may play a role. Specifically, the vertical
shear profile depends on the vertical density stratification
(Eq. \ref{vertical_shear}) which can be enhanced by vertical SG. This
may promote the VSI. Inclusion of the Poisson equation should
therefore be considered.

We have neglected viscosity. \cite{nelson13} have shown that an
alpha viscosity of $\lesssim O(10^{-4})$ is required for the VSI to
operate. It is of interest to quantify the effect of viscosity
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces a meridional
flow in the equilibrium state, as well as viscous heating
in the energy equation.

% \emph{Radial dependence.} Based on previous works, including
% non-linear simulations, which show that the VSI is relevant for
% perturbations with small radial lengthscales, we have adopted the
% radially-local approximation throughout this paper.  Although this
% simplifies the linear problem to ordinary differential equations, this
% approach introduces the wavenumber $k_x$ as a free parameter. The
% effect of radial boundary conditions cannot be explored. Future 
% generalizations of the linear problem should retain the radial
% dependence properly and solve the two-dimensional $(r,z)$ partial 
% differential equation eigenvalue problem.

% Extrapolating our results from the shearing box to a global disk  
% has an inconsistency. Namely, the equilibrium temperature in the
% box only depends on height (if at all). However, it is precisely the
% radial temperature (or entropy) gradient in the global disk that
% gives rise to vertical shear. We do not expect this to be a serious issue for 
% perturbations with small radial wavelengths, 
% but will need to be resolved by adopting global disk models  
% if one wishes to generalize the VSI to large-scale perturbations.

% \emph{Theoretical generalizations.} Our analytical discussion
% (\S\ref{analytical}) can be extended in several ways. One could
% relax the low-frequency approximation to improve the agreement between
% numerical and analytical eigenfrequencies at small $\khat$. However,
% such modes are radially more global, and would need a global
% cylindrical disk model to study them. Another minor improvement is to

%analytical extensions -> nonisothermal, all frequency, non-keplerian,
%diffusion 
%viscosity self-gravity 
%disk models: polytropic connected with isothermal

>>>>

\section{Summary and discussion}\label{summary}
In this paper we have studied the  axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

We generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, by relaxing the Boussinesq and vertically local
approximations adopted in that analysis.

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach allows the thermodynamic
response of the disk to be controlled by varying the parameter
$\beta$, and has been employed in previous numerical simulations of
the VSI \citep{nelson13}. However, it is possible to 
relate  $\beta$ to physical thermal timescales expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}. % who showed the linear VSI studied by
% \cite{nelson13} in a locally isothermal disk can be captured by the
% VGSB. 
We find the VGSB framework is appropriate for non-isothermal
perturbations provided $\beta\lesssim1$.  This is convenient because
rapid thermal relaxation ($\beta\ll1$) is 
generally required for the VSI.  However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$) as 
this is the relevant case for protoplanetary disks that are passively irradiated by its central star 
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation. We obtained explicit solutions describing the VSI with 
thermal relaxation in the thin disk limit.

The fundamental VSI mode is characterized by a symmetric vertical
velocity perturbation with zero nodes and $\delta
v_z\sim\mathrm{constant}$ near the mid-plane, and by an anti-symmetric
density perturbation with one node at $z=0$.  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be the dominant
mode. The fundamental mode is
thus likely the most relevant VSI mode in protoplanetary disks.

We determined the critical thermal relaxation timescale 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the VSI 
operates. We have therefore provided a quantitative measure for the `fast'
thermal relaxation requirement for the VSI in astrophysical disks.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analysis, and accounted for
the background radial disk structure.  Our
numerical results are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.  We find the fundamental mode is robust against
boundary conditions, and higher order modes are more rapidly
stabilized with increasing $\beta$.

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI does not require rapid thermal
relaxation.  When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebulae
(MMSN) described in \cite{chiang10} and   
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity. This results in a $\beta$ function that
depends on both the perturbation lengthscale as well as the height 
from the midplane. We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating the condition
$\beta <\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN. This simple criteron is
thus a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

@@

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $\beta\sim 1$ and disturbances with $|k_x|\ll |k_z|$, where
$k_z$ is a vertical wavenumber. These conditions are mutually exclusive with the VSI. 
The maximum growth rate for convective overstability is given by 
\begin{align}
  \nu_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.

Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
we find  
\begin{align}
  \nu_\mathrm{conv,max} =
  -\frac{\epsilon^2\Omega_k}{4\gamma\Gamma}\left(p+q\right)
  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
\end{align}
% We see that convective overstability is generally slower 
% than the VSI because $\nu_\mathrm{conv,max} = O(\epsilon^2\Omega_k)$,
% whereas the maximum VSI growth rate is $O(\epsilon\Omega_k)$. 
In fact, for the MMSN considered in this work (\S\ref{application} and Appendix
\ref{mmsn}), convective overstability does not apply because
$N_r^2>0$.

However, for more general disk structures convective overstability is
possible. The models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
requires considering perturbations with long radial lengthscales,
$\khat\ll 1$, for which radially global disk models would be more 
appropriate.  
% We will explore convective overstability more fully in a
% follow-up study.

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects
can be represented by a thermal relaxation or cooling timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties. One one can improve the realism further by generalizing the 
opacity model adopted here to be density (and hence height) dependent. 
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment for radiative diffusion. The linear problem then
involves a Laplacian operator, which require additional boundary
conditions. We will consider these improvements in a
follow-up study. %additional heat source

\subsubsection{Additional physics} 
Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a protoplanetary disk, where we estimate the VSI 
operates, disk self-gravity (SG) may play a role. Vertical
self-gravity may favor the VSI by increasing $|\p_z\Omega|$ through
enhancing the density stratification 
(Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

We have neglected viscosity. \cite{nelson13} find that an 
alpha viscosity $\lesssim O(10^{-4})$ is required for the VSI. 
It is of interest to quantify the effect of viscosity 
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces a meridional
flow in the equilibrium state, as well as viscous heating
in the energy equation.
